A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm

### 2014-10-29 paint-contour-tree.tex
making changes to intro

\newcommand{\LL}{\mathbb{L}}
\newcommand{\MM}{\mathbb{M}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\SSS}{\mathbb{S}}

@@

\begin{abstract}
Reeb graphs are a fundamental topological structure which tracks the evolution of levels sets of a real valued function over a manifold.  
In the case when this graph is loop free it is called a contour tree.  Due to the use of the contour trees in practice on large data sets, 
there has been significant previous work in developing algorithms for their efficient computation.

Here we present the first contour tree algorithm which takes into account the shape of the contour tree, 
and in doing so yields instance optimal results.
Specifically, we prove that for any partition of the contour tree into a set of descending paths, $P$, 
any comparison based algorithm requires $\sum_{p\in P} |p|\log|p|$ time.  Moreover, we show there   
exists some partition $P'$ such that the running time of our algorithm is 
$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold 
(and hence this is optimal on all instances requiring $\Omega(N\alpha(N))$ time).
In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, 
while previous algorithms ran in $O(N+t\log t)$ (where $t$ is the size of the contour tree).
%In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, yet still 
%matches the previous worst case $O(N+n\log n)$ bound when the tree is long and skinny.

@@

>>>>

\newcommand{\DD}{\mathbb{D}}
\newcommand{\LL}{\mathbb{L}}
\newcommand{\MM}{\mathbb{M}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\SSS}{\mathbb{S}}

@@

\begin{abstract}
Consider a scalar field $f:\MM \mapsto \RR$, where $\MM$ is a simplicial mesh in $\RR^d$. 
A level set, or contour, at value $v$ is a connected component of $f^{-1}(v)$. 
As $v$ is changed, these contours change topology, merge into each other, or split.
\emph{Contour trees} are concise representations of $f$ that track this contour
behavior. The vertices of these trees are the critical points of $f$, where the gradient is zero.
The edges represent changes in the topology of contours.
It is a fundamental data structure in data analysis and visualization,
and there is much past work (theoretical and practical) on algorithms for construct contour trees.

Suppose $\MM$ has $n$ vertices, $N$ facets, and $t$ critical points.
A classic result of Carr, Snoeyink, and Axen (2000) give an algorithm that takes $O(n\log n + N\alpha(N))$ time
(where $\alpha(\cdot)$ is the inverse Ackermann function). A further improvement to $O(t\log t + N\alpha(N))$ time
was given by Chiang et al. All these algorithms involve a global sort of the critical points,
which is a major bottleneck for large datasets. Unfortunately, lower bounds of $\Omega(t\log t)$ also exist.

We present the first algorithm that can avoid the global sort and has a refined time complexity that
depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
in running time. For a partition of the contour tree into a set of descending paths, $P$,
our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + N\alpha(N))$. If the contour tree is a balanced
binary tree, our running time is $O(n + N\alpha(N))$. Moreover, we prove a lower bound of
exactly $\Omega(\sum_{p \in P} |p|\log |p|)$ for computing contour trees, thereby showing
that the tree structure must play a role in the time complexity.

\ignore{
Moreover, we show there   
exists some partition $P'$ such that the running time of our algorithm is 
$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold

All previous contour tree algorithms involve a global sort 
of the heights of all points or critical vertices. Thereby, these algorithms always pay $\Omega(t \log t)$.
This is a major roadblock for efficiency, since sorting is the most expensive operation

Reeb graphs are a fundamental topological structure which tracks the evolution of levels sets of a real valued function over a manifold.  
In the case when this graph is loop free it is called a contour tree.  Due to the use of the contour trees in practice on large data sets, 
there has been significant previous work in developing algorithms for their efficient computation.

Here we present the first contour tree algorithm which takes into account the shape of the contour tree, 
and in doing so yields instance optimal results.
Specifically, we prove that for any partition of the contour tree into a set of descending paths, $P$, 
any comparison based algorithm requires $\sum_{p\in P} |p|\log|p|$ time.  Moreover, we show there   
exists some partition $P'$ such that the running time of our algorithm is 
$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold 
(and hence this is optimal on all instances requiring $\Omega(N\alpha(N))$ time).
In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, 
while previous algorithms ran in $O(N+t\log t)$ (where $t$ is the size of the contour tree).
%In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, yet still 
%matches the previous worst case $O(N+n\log n)$ bound when the tree is long and skinny.
}

@@

A large class of data, such as satellite terrain maps, physical simulations, combustion data, can
be thought as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
by consider $f$ to be piecewise linear over some mesh (simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a specific topological data structure used to represent and visualize
the function $f$.

It is convenient to think of $f$ as a $(d+1)$-dimensional manifold, or alternately consider
$\MM$ in $\RR^{d+1}$ with the last coordinate (height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. At the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other,
split into new components, contours acquire or lose handles. These are the relevant events where 
the topology of the level set changes. The contour tree is a concise representation of all these events.

For convenience, imagine $f$ to be smooth. All points where the gradient of $f$ is zero are \emph{critical point},
exactly where the topology of the contours change. The vertices of the contour tree are exactly the critical points.

Contour trees and related topological data structures have been applied in analysis of fluid mixing, combustion simulations,
and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}.
